Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-05 submission.tex
added mft to results

\section{Previous Work}
\label{sec:prev}
There has been little previous work dedicated specifically to
extrinsic evaluation of the effects of PoS tag sets on downstream
applications.  There has, however, been some work that evaluates the
effects of tag set granularity on PoS tagging itself. This includes
investigation of the effects of tag sets on PoS tagging for Swedish
\cite{Meg:01,Meg:02} and English \cite{Mac:05}.

@@

\newcite{Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a tag set based
on a Swedish version of PAROLE tags totaling 139 tags. Furthermore, the effects 
of tag set size on tagging was investigated by mapping the original tag
set into smaller subsets designed for parsing. \newcite{Meg:02} argues that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. The study found that the smallest tag set comprising 26
tags yields the lowest tagger error rate.  However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. \newcite{Meg:02} argues that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance.  Unfortunately, results for parsing with the various PoS
tag sets are not reported.

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of the Penn
Treebank (PTB) \cite{Mar:San:Mar:93}. Based on linguistic
considerations \newcite{Mac:05} mapped the 45 tags of the original PTB tag set
 to more fine-grained tag sets to investigate whether additional linguistic information 
could assist the tagger. Experimenting with both lexically
and syntactically conditioned modifications such as distinguishing between
count nouns and noncount nouns and between transitive and intransitive verbs,
they found that more fine-grained tag sets rarely led to improvements in tagger
accuracy; the most successful modification yielded an improvement in tagger
accuracy of 0.05 percentage points.  They did not find any statistically
significant improvements using linguistically informed distinctions, arguing
that their results do not support the hypothesis that it is possible to achieve
significant performance improvements in PoS tagging by utilizing a
finer-grained tag set.

% fixme:
Transition/overgang her ...

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity. To evaluate the initial tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we ran
baseline experiments with the two tag sets. In Table \ref{inittagseteval}, we
report the results of these experiments. We see that the tagger accuracy
drastically drops when going from the original to the full tag set. TnT reports
an accuracy of 97.47\% on the original tag set, which is reduced to 93.48\% for
the full tag set. These results confirm our hypothesis that the very high
linguistic quality in the full, fine-grained tag set comes at the expense of
drops in tagger performance.  However, the additional linguistic information
provided by the full tag set improves the parser performance. With the original
tag set, Mate reports an LAS of 87.01\% and a UAS of 90.19\%, which increases
to 87.15\% and 90.39\%, respectively, using the full tag set. As we are looking
for linguistically informed distinctions that improve the syntactic parsing,
these results are promising and serve to indicate that additional linguistic
information assists the syntactic parsers, motivating the optimization of the
existing PoS tag set.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Tag Set} & \textbf{Acc} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 93.48\% & \textbf{87.15\%} & \textbf{90.39\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the two initial tag sets.}
    \label{inittagseteval}
\end{table}

@@

\subsection{Optimized Tag Set}
\begin{table*}[ht]
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Category} & \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} &
        \textbf{UAS} \\
        \midrule
        \emph{Baseline} & --- & 97.47\% & 87.01\% & 90.19\% \\
        Noun & Type, case \& definiteness & 97.05\% & 88.81\% & 91.73\% \\
        Verb & Finiteness & 97.35\% & 87.30\% & 90.43\% \\
        Adjective & Degree & 97.41\% & 87.29\% & 90.44\% \\
        Determiner & Definiteness & 97.49\% & 87.30\% & 90.42\% \\
        Pronoun & Type \& case & 97.51\% & 87.30\% & 90.41\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the most successful tag set
        modification for each category.}
    \label{respectiveresults}
\end{table*}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Data} & \textbf{Tag Set} & \textbf{Acc}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{97.30\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}

>>>>

\section{Previous Work}
\label{sec:prev}
There has been little previous work dedicated specifically to
extrinsic evaluation of the effects of PoS tag sets on downstream
applications.  There has, however, been some work that evaluates the
effects of PoS tag set granularity on PoS tagging. This includes
investigation of the effects of PoS tag sets on tagging of Swedish
\cite{Meg:01,Meg:02} and English \cite{Mac:05}.

@@

\newcite{Meg:02} trained and evaluated a range of PoS taggers on the
Stockholm-Umeå Corpus (SUC) \cite{Gus:Har:06}, annotated with a tag set based
on a Swedish version of PAROLE tags totaling 139 tags. Furthermore, they
investigated the effects of tag set size on tagging by mapping the original tag
set into smaller subsets designed for parsing. They argue that a tag set with
complete morphological tags may not be necessary for all NLP applications, for
instance syntactic parsing. They found that the smallest tag set comprising 26
tags yields the lowest tagger error rate.  However, for some of the taggers,
augmenting the tag set with more linguistically informative tags may actually
lead to a drop in error rate. They argue that this shows that the size of the
tag set as well as the type of information in the tags are crucial factors for
tagger performance.  However/unfortunately, they do not report results of
parsing with the various PoS tag sets.

Similarly, \newcite{Mac:05} investigated the effects of PoS tag sets on tagger
performance in English, specifically the Wall Street Journal portion of Penn
Treebank \cite{Mar:San:Mar:93}. They mapped the original tag set of Penn
Treebank comprising 45 tags to more fine-grained tag sets using linguistic
insight to investigate whether additional linguistic information included in
finer-grained tags could assist the tagger. Experimenting with both lexically
and syntactically conditioned modifications such as distinguishing between
count nouns and noncount nouns and between transitive and intransitive verbs,
they found that more fine-grained tag sets rarely led to improvements in tagger
accuracy; the most successful modification yielded an improvement in tagger
accuracy of 0.05 percentage points.  They did not find any statistically
significant improvements using linguistically informed distinctions, arguing
that their results do not support the hypothesis that it is possible to achieve
significant performance improvements in PoS tagging by utilizing a
finer-grained tag set.

Transition/overgang her ...

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity. To evaluate the initial tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we ran
baseline experiments with the two tag sets. In Table \ref{inittagseteval}, we
report the results of these experiments. We see that the tagger accuracy
drastically drops when going from the original to the full tag set. The MFT
baseline for the original tag set is 94.14\%, while it drops by almost 9
percentage points to 85.15\% for the full tag set. TnT reports an accuracy of
97.47\% on the original tag set, which is reduced to 93.48\% for the full tag
set. These results confirm our hypothesis that the very high linguistic quality
in the full, fine-grained tag set comes at the expense of drops in tagger
performance.  However, the additional linguistic information provided by the
full tag set improves the parser performance. With the original tag set, Mate
reports an LAS of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and
90.39\%, respectively, using the full tag set. As we are looking for
linguistically informed distinctions that improve the syntactic parsing, these
results are promising and serve to indicate that additional linguistic
information assists the syntactic parsers, motivating the optimization of the
existing PoS tag set.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Tag Set} & \textbf{MFT} & \textbf{Acc} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 85.15\% & 93.48\% & \textbf{87.15\%} & \textbf{90.39\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the two initial tag sets.}
    \label{inittagseteval}
\end{table}

@@

\subsection{Optimized Tag Set}
\begin{table*}[ht]
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Category} & \textbf{Feature(s)} & \textbf{MFT} &
        \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \emph{Original} & --- & 94.14\% & 97.47\% & 87.01\% & 90.19\% \\
        Noun & Type, case \& definiteness & 89.61\% & 97.05\% & 88.81\% &
        91.73\% \\
        Verb & Finiteness & 93.72\% & 97.35\% & 87.30\% & 90.43\% \\
        Adjective & Degree & 94.13\% & 97.41\% & 87.29\% & 90.44\% \\
        Determiner & Definiteness & 94.13\% & 97.49\% & 87.30\% & 90.42\% \\
        Pronoun & Type \& case & 94.12\% & 97.51\% & 87.30\% & 90.41\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the most successful tag set
        modification for each category.}
    \label{respectiveresults}
\end{table*}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Acc} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 85.15\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.30\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}
